Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-26 submission/manuscript.Rmd
edit discussion 3

Dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis, as germline and other tissue-enriched genes are aberrantly expressed in epiblast-like cells (EpiLCs). _In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into the body's somatic tissues[@bardotMouseGastrulationCoordination2020]. However, a small subset of epiblast stem cells will receive signals to reactivate germline gene expression to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011]. Thus, misexpression of germline genes in _Kdm5c_-KO EpiLCs might suggest they are progressing beyond EpiLC differentiation and becoming PGCLCs. Yet, _Kdm5c_-KO EpiLCs had proper expression of EpiLC marker genes for primed pluripotency and we observed no difference in cellular morphology during ESC to EpiLC differentiation. Furthermore, we saw no significant change in _Otx2_ expression, an epiblast stem cell marker that is known to repress epiblast and EpiLC differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018]. This, in conjunction with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to germline identity.

- change first sentence: Other chromatin regulators involved in germline gene repression act in ESCs and EpiLCs. We observed this is true for KDM5C

>>>>

RNA-seq of epiblast-like cells (EpiLCs) revealed dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis. _In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into the body's somatic tissues[@bardotMouseGastrulationCoordination2020]. However, a small subset of epiblast stem cells will receive signals to reactivate germline gene expression to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011]. Therefore, misexpression of germline genes in EpiLCs might suggest _Kdm5c_-KO ESCs are progressing beyond EpiLC differentiation and instead becoming PGCLCs. Yet, _Kdm5c_-KO EpiLCs properly express primed pluripotency genes, including _Otx2_ - an epiblast stem cell marker that is known to repress epiblast and EpiLC differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018]. Furthermore, we observed no difference in cellular morphology during _Kdm5c_-KO ESC to EpiLC differentiation. Proper EpiLC differentiation, together with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.
